---
title: "Class12: Population Scale Analysis"
author: "Joshua Chun (A17812847)"
format: pdf
toc: true
---

```{r}
# Read genotype and expression data from the class website
expr <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt",
                   header = TRUE)

# Check the first few rows
head(expr)

# Check column names
colnames(expr)
```

```{r}
# Count sample size for each genotype
table(expr$geno)
```

```{r}
# Calculate median expression for each genotype
tapply(expr$exp, expr$geno, median)
```


> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The number of samples for each genotype group was:
- A/A: 108 samples
- A/G: 233 samples
- G/G: 121 samples

The median ORMDL3 expression values for each genotype were:
- A/A: 31.24847
- A/G: 25.06486
- G/G: 20.07363

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
# Make a boxplot of ORMDL3 expression by genotype
boxplot(exp ~ geno,
        data = expr,
        xlab = "Genotype",
        ylab = "ORMDL3 Expression",
        main = "ORMDL3 Expression by rs8067378 Genotype")
```

The boxplot shows that ORMDL3 expression differs among rs8067378 genotype groups. Individuals with the A/A genotype show the highest median ORMDL3 expression, while individuals with the G/G genotype show the lowest expression. This shows that variation at rs8067378 may influence ORMDL3 gene expression and could contribute to asthma susceptibility.