---
title: "class08"
author: "Joshua Chun (A17812847)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. 

Today we will analyze a biopsy data-set from fine needle aspiration (FNA) of a breast mass.

## Data Import

The data is made available as a CSV file for download. We can read this using `read.csv()`:

```{r}
# Read the Wisconsin cancer dataset from a CSV file
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

```{r}
# Display the first 3 rows of the dataset
head(wisc.df, 3)
```

Make sure we remove or exclude the `diagnosis` column from the data-set that we use for further analysis - this is the expert diagnosis as either M or B:

```{r}
# We can use -1 here to remove the first column and create diagnosis vector for later
wisc.data <- wisc.df[,-1]
diagnosis <- as.factor(wisc.df$diagnosis)
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
# Find number of rows (observations/samples)
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
# Count how many are Malignant (M) vs Benign (B)
table(wisc.df$diagnosis)
```

```{r}
# Count how many are Malignant (M)
sum(wisc.df$diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with _mean?

We can use the `grep()` function to help us here:

```{r}
# Return all the column names
colnames(wisc.data)
```

```{r}
# Find column names that contain "_mean" and count matches found
length(grep("_mean", colnames(wisc.data), value = T))
```

## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale=TRUE` argument to `prcomp()`.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Let's see the "PC score plot"
```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

```{r}
summary(wisc.pr)$importance[2,1]
```

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
# Extract the cumulative proportion of variance explained by each principal component
cumvar <- summary(wisc.pr)$importance[3,]

# Find the first principal component where cumulative variance is at least 70%
which(cumvar >= 0.70)[1]
```

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

```{r}
# Extract the cumulative proportion of variance explained by each principal component
cumvar <- summary(wisc.pr)$importance[3,]

# Find the first principal component where cumulative variance is at least 90%
which(cumvar >= 0.90)[1]
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

```{r}
# Plot using coordinate data, plot as a scatter plot
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

The biplot is difficult to understand because it is crowded with too many observation labels and variable arrows all plotted at once. This makes it hard to see patterns or separation between groups. A simple scatter plot of PC1 vs PC2 is much easier to interpret.

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
# Repeat for components 1 and 3
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```

The PC1 vs PC3 plot shows much weaker separation between malignant and benign samples compared to the PC1 vs PC2 plot. PC1 still separates the groups, while PC3 does not provide much additional distinction, and the points appear more mixed.

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
# Get loading value for concave.points_mean on PC1
wisc.pr$rotation["concave.points_mean", 1]
```

The loading value for concave.points_mean on PC1 is large, which shows it strongly contributes to the first principal component. However, there are other features (such as radius_mean, perimeter_mean, and area_mean) that have similar or slightly larger contributions.

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
# Calculate Euclidean distance between all observations
data.dist <- dist(data.scaled)
```

```{r}
# Perform hierarchical clustering using complete linkage
wisc.hclust <- hclust(data.dist, method = "complete")
```

```{r}
# Plot dendrogram
plot(wisc.hclust)

# Draw horizontal line where ~4 clusters appear
abline(h = 19, col = "red", lty = 2)
```

```{r}
# Cut the dendrogram into 4 clusters
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
# Compare clusters to true labels
table(wisc.hclust.clusters, diagnosis)
```
The clustering model has approximately 4 clusters at a height of around 19–20.


> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
# Try different hierarchical clustering methods
hc.single <- hclust(data.dist, method = "single")
hc.complete <- hclust(data.dist, method = "complete")
hc.average <- hclust(data.dist, method = "average")
hc.ward <- hclust(data.dist, method = "ward.D2")

# Compare how each method separates diagnosis using 4 clusters
table(cutree(hc.single, k = 4), diagnosis)
table(cutree(hc.complete, k = 4), diagnosis)
table(cutree(hc.average, k = 4), diagnosis)
table(cutree(hc.ward, k = 4), diagnosis)
```

My favorite method is ward.D2 because it gives more balanced and biologically meaningful clusters compared to the other methods. Since ward.D2 minimizes the variance within clusters, it tends to group samples that are more similar overall. This makes sense for this data set because malignant and benign samples should cluster based on similar nucleus measurements. Overall, it gives clearer separation between malignant and benign diagnoses than methods like single linkage, which can create messier chained clusters.

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
# 1. Create clustering model using first 7 PCs
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method = "ward.D2")

# 2. Cut into 2 clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)

# 3. Now compare to diagnosis
table(wisc.pr.hclust.clusters, diagnosis)

```

The PCA-based hierarchical clustering separates the data very well into two clusters. One cluster contains mostly malignant samples (188 M vs 28 B), while the other contains mostly benign samples (329 B vs 24 M). This indicates strong separation with relatively few mismatches. 

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```


The hierarchical clustering without PCA performs worse at separating the diagnoses. The clusters are less cleanly divided, with more mixing of malignant and benign samples across multiple clusters. Compared to the PCA-based clustering, this is less effective at clearly distinguishing between the two groups.

> Q16. Which of these new patients should we prioritize for follow up based on your results?

```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
# Read in new patient samples
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)

# Project new samples onto the PCA model
npc <- predict(wisc.pr, newdata = new)

# Plot original samples using PC1 and PC2
plot(wisc.pr$x[,1:2], col = grps)

# Add new patient samples as large blue points
points(npc[,1], npc[,2], col = "blue", pch = 16, cex = 3)

# Label the two new samples
text(npc[,1], npc[,2], c(1, 2), col = "white")
```

Patient 2 should be prioritized for follow-up because it clusters with the malignant samples in PCA space, while Patient 1 clusters with the benign group.
