---
title: "class14"
author: "Joshua Chun (A17812847)"
format: pdf
toc: true
---

# Differential Expression Analysis

```{r}
library(DESeq2)

metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata
colData <- read.csv(metaFile, row.names = 1)
head(colData)

# Import count data
countData <- read.csv(countFile, row.names = 1)
head(countData)
```

> Q1. Remove the troublesome first column from countData

```{r}
# The first column is "length", which is not a sample count column
# We remove column 1 and keep only the sample count columns
countData <- as.matrix(countData[, -1])

# Check that the length column is gone
head(countData)
```

> Q2. Filter out genes with zero reads across all samples

```{r}
# rowSums(countData) adds the counts across all samples for each gene
# We keep only genes where the total count is greater than 0
countData <- countData[rowSums(countData) > 0, ]

# Check filtered count data
head(countData)
```

# Run DESeq2

```{r}
# Create DESeq2 dataset object
# countData is the gene count matrix
# colData is the metadata
# design is the condition being tested
dds <- DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ condition
)

# Run DESeq2 pipeline
dds <- DESeq(dds)
dds
```

# Get DESeq2 results

```{r}
# Get differential expression results
res <- results(dds)

# Check result names
resultsNames(dds)

# Preview results
head(res)
```

> Q3. Summarize DESeq2 results

```{r}

# Summarize number of upregulated and downregulated genes
summary(res)
```

# Volcano Plot

```{r}
# Load ggplot2 for plotting
library(ggplot2)

# Basic volcano plot:
# x-axis = log2 fold change
# y-axis = -log adjusted p-value
ggplot(as.data.frame(res)) +
  aes(
    x = log2FoldChange,
    y = -log(padj)
  ) +
  geom_point()
```

> Q4. Improved volcano plot with colors and cutoff lines

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res))

# Color blue genes with fold change above 2
mycols[abs(res$log2FoldChange) > 2] <- "blue"

# Color gray genes with adjusted p-value greater than 0.01
mycols[res$padj > 0.01] <- "gray"

# Make volcano plot
ggplot(as.data.frame(res)) +
  aes(
    x = log2FoldChange,
    y = -log(padj)
  ) +
  geom_point(color = mycols) +
  xlab("Log2(FoldChange)") +
  ylab("-Log(P-value)") +
  geom_vline(xintercept = c(-2, 2)) +
  geom_hline(yintercept = -log(0.01))
```


```{r}
# Create logical vector for significant genes
sig <- res$padj < 0.05 &
       abs(res$log2FoldChange) > 2

# Make volcano plot highlighting significant genes
ggplot(as.data.frame(res)) +
  aes(
    x = log2FoldChange,
    y = -log(padj),
    color = sig
  ) +
  geom_point() +
  xlab("Log2(FoldChange)") +
  ylab("-Log(P-value)") +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log(0.05),
    linetype = "dashed"
  ) +
  theme_bw()

```

# Add gene annotation

```{r}
# Load annotation packages
library(AnnotationDbi)
library(org.Hs.eg.db)

# Show available annotation columns
columns(org.Hs.eg.db)
```

> Q5. Add SYMBOL, ENTREZID, and GENENAME annotations

```{r}

# Add gene symbols
res$symbol <- mapIds(
  org.Hs.eg.db,
  keys = row.names(res),
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)

# Add Entrez IDs
res$entrez <- mapIds(
  org.Hs.eg.db,
  keys = row.names(res),
  keytype = "ENSEMBL",
  column = "ENTREZID",
  multiVals = "first"
)

# Add full gene names
res$name <- mapIds(
  org.Hs.eg.db,
  keys = row.names(res),
  keytype = "ENSEMBL",
  column = "GENENAME",
  multiVals = "first"
)

# Preview annotated results
head(res, 10)
```

> Q6. Reorder results and save to CSV

```{r}
# Order results by adjusted p-value
res <- res[order(res$padj), ]

# Save results to CSV file
write.csv(res, file = "deseq_results.csv")
```

# Section 2: Pathway Analysis

```{r}
# Load pathway analysis packages
library(pathview)
library(gage)
library(gageData)

# Load KEGG pathway gene sets
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs <- kegg.sets.hs[sigmet.idx.hs]

# Examine first 3 pathways
head(kegg.sets.hs, 3)
```

# Create named fold change vector for GAGE

```{r}
# Extract log2 fold changes
foldchanges <- res$log2FoldChange

# Name fold changes with Entrez gene IDs
names(foldchanges) <- res$entrez

# Preview fold change vector
head(foldchanges)
```

# Run KEGG pathway analysis

```{r}
# Run GAGE pathway analysis using KEGG gene sets
keggres <- gage(
  foldchanges,
  gsets = kegg.sets.hs
)

# Check structure of results
attributes(keggres)

# View top upregulated pathways
head(keggres$greater)

# View top downregulated pathways
head(keggres$less)
```

# Pathview example: Cell cycle pathway

```{r}
# Plot KEGG cell cycle pathway
pathview(
  gene.data = foldchanges,
  pathway.id = "hsa04110"
)

# Alternative PDF-style output
pathview(
  gene.data = foldchanges,
  pathway.id = "hsa04110",
  kegg.native = FALSE
)
```

# Top 5 upregulated KEGG pathways

```{r}
# Get names of top 5 upregulated pathways
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract KEGG pathway IDs
keggresids <- substr(
  keggrespathways,
  start = 1,
  stop = 8
)

# View pathway IDs
keggresids

# Plot top 5 upregulated pathways
pathview(
  gene.data = foldchanges,
  pathway.id = keggresids,
  species = "hsa"
)
```


> Q7: Plot top 5 downregulated KEGG pathways

```{r}
# Get names of top 5 downregulated pathways
down.pathways <- rownames(keggres$less)[1:5]

# Extract KEGG pathway IDs
down.ids <- substr(
  down.pathways,
  start = 1,
  stop = 8
)

# View downregulated pathway IDs
down.ids

# Plot top 5 downregulated pathways
pathview(
  gene.data = foldchanges,
  pathway.id = down.ids,
  species = "hsa"
)
```

# Section 3: Gene Ontology Analysis

```{r}
# Load GO gene sets
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process GO terms
gobpsets <- go.sets.hs[go.subs.hs$BP]

# Run GAGE on GO biological process terms
gobpres <- gage(
  foldchanges,
  gsets = gobpsets
)

# View top GO results
lapply(gobpres, head)
```

# Section 4: Reactome Analysis

```{r}
# Extract significant genes with adjusted p-value <= 0.05
sig_genes <- res[
  res$padj <= 0.05 & !is.na(res$padj),
  "symbol"
]

# Print total number of significant genes
print(
  paste(
    "Total number of significant genes:",
    length(sig_genes)
  )
)

# Save significant genes as a plain text file
write.table(
  sig_genes,
  file = "significant_genes.txt",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)
```

> Q. What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The pathway with the most significant “Entities p-value” in the analysis was related to the Cell Cycle/Mitotic Cell Cycle pathway. These results were consistent with the previous KEGG pathway analysis, since the Cell Cycle pathway was also one of the most significantly downregulated pathways. Other related pathways, such as DNA replication and mitosis, were also identified in both analyses. The results between Reactome and KEGG were similar overall because both analyses detected strong disruption of genes involved in cell division and cell cycle regulation after HOXA1 knockdown. However, some differences occurred because Reactome and KEGG use different pathway databases, pathway definitions, and statistical methods. Also, differences in gene annotations, pathway sizes, and significance thresholds can lead to variation in which pathways appear most enriched.



