---
title: "HW6 R Functions"
author: "Joshua Chun (A17812847)"
output: html_document
date: "2026-04-17"
---

## Q1

The type of object returned from read.pdb() is a list object, which returns a list containing components like atom data, coordinates, etc.

## Q2

This function extracts a subset of the PDB structure, like selecting chain A or selecting alpha carbons only. For example, chain=”A” selects chain A, or doing elety=”CA” selects alpha carbons only.

## Q3

The parameter sse = NULL will turn off the rectangles in the plots. They represent the secondary structure elements like the alpha helices and the beta sheets.

## Q4

An overlay plot (single combined line plot) would be a better plot to compare proteins. We could plot all the B-factors and use different colors for a direct comparison across proteins.

## Q5

The proteins s2 and s3 are the most similar because they cluster together at the lowest height in the dendrogram. This indicates they have the smallest distance between their B-factor profiles. Similarity is shown using a distance matrix (dist) followed by hierarchical clustering (hclust), where lower linkage height shows greater similarity.

## Q6

```{r}
library(bio3d)
```

```{r}
s1 <- read.pdb("4AKE")
s2 <- read.pdb("1AKE")
s3 <- read.pdb("1E4Y")

s1.chainA <- trim.pdb(s1, chain = "A", elety = "CA")
s2.chainA <- trim.pdb(s2, chain = "A", elety = "CA")
s3.chainA <- trim.pdb(s3, chain = "A", elety = "CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
```

```{r}
hc <- hclust(dist(rbind(s1.b, s2.b, s3.b)))
plot(hc)
```


```{r}
analyze_protein <- function(pdb_id) {
  pdb <- read.pdb(pdb_id)
  protein <- trim.pdb(pdb, chain = "A", elety = "CA")
  b <- protein$atom$b
  
# Input:
# pdb_id = a PDB ID such as "4AKE"
  
# What this function does: 
# This function reads a protein PDB structure, selects chain A CA atoms, extracts the B-factors, and makes a B-factor plot.

# Output:
# Returns a vector of B-factor values and displays a plot.
  
  plotb3(b, sse = protein, typ = "l", ylab = "Bfactor",
         main = paste("B-factor plot for", pdb_id))
  
  return(b)
}
```

```{r}
analyze_protein("4AKE")
```

